---
title: "LV Strain Echo"
author: "Kipp Johnson"
date: "November 29, 2015"
output: 
  html_document: 
    number_sections: yes
    toc: yes
---

```{r Preliminaries, echo=FALSE}
rm(list=ls())
```

# R function to read echo files

R function called read.echo() to parse the input echo files and extract:
* time
* strain
* strain rate
* radial strain
* radial strain rate
* volume
* segment volume

```{r Defining read.echo()}

read.echo <- function(infile){

  # Read in the file into a single vector
  con <- readLines(infile)

  # Instantiate vectors
  frame.time <- vector()
  time.progression <- vector()
  strain <- vector()
  strain.rate <- vector()
  radial.strain <- vector()
  radial.strain.rate <- vector()
  volume <- vector()
  seg.vol <- vector()

  for(i in 1:length(con)){
    if(con[i]=="FrameTime(ms)"){
      frame.time <- strsplit(con[i+1], split="\t")[[1]]
      }

    if(con[i]=="TimeProgression(ms)"){
      time.progression <- strsplit(con[i+1], split="\t")[[1]]
      }

    if(con[i]=="Strain (%)"){
      j <- 1
      while(con[i+j] != ""){
        strain <- rbind(strain, strsplit(con[i+j], split="\t")[[1]])
        j <- j + 1
      }
    }

    if(con[i]=="StrainRate (1/s)"){
      j <- 1
      while(con[i+j] != ""){
        strain.rate <- rbind(strain.rate, strsplit(con[i+j], split="\t")[[1]])
        j <- j + 1
      }
    }

    if(con[i]=="RadialStrain (%)"){
      j <- 1
      while(con[i+j] != ""){
        radial.strain <- rbind(radial.strain, strsplit(con[i+j], split="\t")[[1]])
        j <- j + 1
      }
    }

    if(con[i]=="RadialStrainRate (1/s)"){
      j <- 1
      while(con[i+j] != ""){
        radial.strain.rate <- rbind(radial.strain.rate, strsplit(con[i+j], split="\t")[[1]])
        j <- j + 1
      }
    }

    if(con[i]=="Vol (ml)"){
      volume <- strsplit(con[i+1], split="\t")[[1]]
      }

    if(con[i]=="Seg_Vol (ml)"){
      j <- 1
      while(con[i+j] != ""){
        seg.vol <- rbind(seg.vol, strsplit(con[i+j], split="\t")[[1]])
        j <- j + 1
      }
    }


    }

  class(frame.time) <- "numeric"
  class(time.progression) <- "numeric"
  class(strain) <- "numeric"
  class(strain.rate) <- "numeric"
  class(radial.strain) <- "numeric"
  class(radial.strain.rate) <- "numeric"
  class(volume) <- "numeric"
  class(seg.vol) <- "numeric"

  return(list(frame.time=(frame.time),
              time.progression=(time.progression),
              strain=t(strain),
              strain.rate=t(strain.rate),
              radial.strain=t(radial.strain),
              radial.strain.rate=t(radial.strain.rate),
              volume=(volume),
              seg.vol=(seg.vol)))
}
```

```{r, Reading in the files}

# Read in the echos
echo1 <- read.echo("~/Projects/partho/data/pborsuk2c.txt")
echo2 <- read.echo("~/Projects/partho/data/pborsuk4c.txt")

# Get number of time points
nTimepoints <- dim(echo1$strain)[1]
nProbes <- dim(echo1$strain)[2]
  
```

# Exploratory Analysis

## Strain

```{r Plot strains, echo=FALSE}
### Strain
# Plot echo #1 strain
plot(echo1$time.progression, echo1$strain[,1], type="l", 
     ylab="Strain (%)", xlab="Time (ms)", main="Sample Echo 1",
     ylim=c( min(echo1$strain), max(echo1$strain)))

for(i in 2:nProbes){
  lines(echo1$time.progression, echo1$strain[,i], type="l", col=i+1)
}

# Plot echo #2 strain
plot(echo2$time.progression, echo2$strain[,1], type="l", 
     ylab="Strain (%)", xlab="Time (ms)", main="Sample Echo 2",
     ylim=c( min(echo2$strain), max(echo2$strain)))

for(i in 2:nProbes){
  lines(echo2$time.progression, echo2$strain[,i], type="l", col=i+1)
}

```

## Strain Rates

```{r Plot strain rates, echo=FALSE}
### Strain Rates
# Plot echo #1 strain rate
plot(echo1$time.progression, echo1$strain.rate[,1], type="l", 
     ylab="Strain Rate (1/s)", xlab="Time (ms)", main="Sample Echo 1",
     ylim=c( min(echo1$strain.rate), max(echo1$strain.rate)))

for(i in 2:nProbes){
  lines(echo1$time.progression, echo1$strain.rate[,i], type="l", col=i+1)
}


# Plot echo #2 strain rate
plot(echo2$time.progression, echo2$strain.rate[,1], type="l",
     ylab="Strain Rate (1/s)", xlab="Time (ms)", main="Sample Echo 2",
     ylim=c( min(echo2$strain.rate), max(echo2$strain.rate)))

for(i in 2:nProbes){
  lines(echo2$time.progression, echo2$strain.rate[,i], type="l", col=i+1)
}

```

## Radial Strain

```{r Radial strains, echo=FALSE}
### Radial strain
# Plot echo #1 radial strain
plot(echo1$time.progression, echo1$radial.strain[,1], type="l", 
     ylab="Radial Strain (%)", xlab="Time (ms)", main="Sample Echo 1",
     ylim=c( min(echo1$radial.strain), max(echo1$radial.strain)))

for(i in 2:nProbes){
  lines(echo1$time.progression, echo1$radial.strain[,i], type="l", col=i+1)
}

# Plot echo #2 radial strain
plot(echo2$time.progression, echo2$radial.strain[,1], type="l", 
     ylab="Radial Strain (%)", xlab="Time (ms)", main="Sample Echo 2",
     ylim=c( min(echo2$radial.strain), max(echo2$radial.strain)))

for(i in 2:nProbes){
  lines(echo2$time.progression, echo2$radial.strain[,i], type="l", col=i+1)
}

```

## Radial strain rates

```{r Radial strain rates, echo=FALSE}

### Radial Strain Rate
# Plot echo #1 radial strain rate
plot(echo1$time.progression, echo1$radial.strain.rate[,1], type="l", 
     ylab="Radial Strain Rate (1/s)", xlab="Time (ms)", main="Sample Echo 1",
     ylim=c( min(echo1$radial.strain.rate), max(echo1$radial.strain.rate)))

for(i in 2:nProbes){
  lines(echo1$time.progression, echo1$radial.strain.rate[,i], type="l", col=i+1)
}

# Plot echo #2 radial strain
plot(echo2$time.progression, echo2$radial.strain.rate[,1], type="l",
     ylab="Radial Strain Rate (1/s)", xlab="Time (ms)", main="Sample Echo 2",
     ylim=c( min(echo2$radial.strain.rate), max(echo2$radial.strain.rate)))

for(i in 2:nProbes){
  lines(echo2$time.progression, echo2$radial.strain.rate[,i], type="l", col=i+1)
}


```

# Generating sample data

Generating by using multivariate normal distribution with means and covariance matrix from given time series.

## Example: plots of randomly generated sample data
```{r, Generate sample data to plot}

library(MASS)
set.seed(10)


# From first echo example
r1 <- mvrnorm(nProbes, mu=apply(echo1$strain, 1, mean), Sigma=cov(t(echo1$strain)))
plot(x=echo1$time.progression, y=r1[1,], ylim=c(min(r1),max(r1)),type="n", xlab="Time (ms)", ylab="Strain", main="Random Echo Generated from Echo 1")
apply(r1, 1, lines, x=echo1$time.progression)


# From second echo example
r2 <- mvrnorm(nProbes, mu=apply(echo2$strain, 1, mean), Sigma=cov(t(echo2$strain)))
plot(x=echo2$time.progression, y=r2[1,], ylim=c(min(r2),max(r2)),type="n", xlab="Time (ms)", ylab="Strain",main="Random Echo Generated from Echo 2")
apply(r2, 1, lines, x=echo2$time.progression)

```

### Dealing with different times
The first echo has time data for 92 time points; the second has data for 93 time points. We can drop the last time point to facilitate analysis, if required later.

```{r Dropping time point}
#plot(echo1$time.progression/echo2$time.progression[1:length(echo1$time.progression)])

t1 <- echo1$time.progression
t2 <- echo2$time.progression[1:length(echo1$time.progression)] #drop last time

```

# Statistical Analysis

# Trace of covariance matrix

The trace of the covariance matrix is equal to the total variance. 

```{r Trace of covariance matrix}
library(psych)

st1 <- echo1$strain
st2 <- echo2$strain

tr(cov(st1))
tr(cov(st2))

```


# Tensor Analysis

## Create tensors of sample data

Uses the rTensor package.

We create multidimensional arrays and then fill them in with simulated data with the following structure:

i = sample index (1 to n samples)
j = index of LV point where strain is measured
k = time

TODO: Make sure R x C are in correct order.

```{r, Actually generate sample data tensors}

nSamples <- 25

# Set dimensions for an i by j by k tensor
# i = strain probe index
# j = time point index
# k = sample index 
tensorDim1 <- c(dim(echo1$strain)[2], dim(echo1$strain)[1], nSamples) 
samp1 <- array(data=NA, dim=tensorDim1)

for(i in 1:nSamples){ 
  samp1[, ,i] <- mvrnorm(nProbes, mu=apply(echo1$strain, 1, mean), Sigma=cov(t(echo1$strain)))
}

#Plot simulated data
# Probe 1, 25 individuals
plot(samp1[1,,1],type="l", ylim=c(min(samp1[1,,]),max(samp1[1,,]) ))
for(i in 2:nSamples){ lines(samp1[1,,i])}

# Probe 2, 25 individuals
plot(samp1[2,,1],type="l", ylim=c(min(samp1[2,,]),max(samp1[2,,]) ))
for(i in 2:nSamples){ lines(samp1[2,,i])}

# Creating tensor 2
tensorDim2 <- c(dim(echo2$strain)[2], dim(echo2$strain)[1], nSamples) 
samp2 <- array(data=NA, dim=tensorDim2)

for(i in 1:nSamples){ 
  samp2[, ,i] <- mvrnorm(nProbes, mu=apply(echo2$strain, 1, mean), Sigma=cov(t(echo2$strain)))
}

library(rTensor)
tnsr1 <- as.tensor(samp1)
tnsr2 <- as.tensor(samp2)
```

# Principal component analysis, strain, actual data

```{r Principal component analysis, strain}

ep1 <- prcomp(echo1$strain, scale=TRUE) # 70% by pc1
ep2 <- prcomp(echo2$strain, scale=TRUE) # 70% by pc1

ep1 <- prcomp(echo1$strain, scale=FALSE) # 70% by pc1
ep2 <- prcomp(echo2$strain, scale=FALSE) # 70% by pc1

#summary(ep1)
#summary(ep2)

#plot(ep1$rotation[,1], ep1$rotation[,2])
#plot(ep2$rotation[,1], ep2$rotation[,2])

plot(ep1$sdev, type="b",xlab="Principal component",ylab="Variance",main="Echo 1")
plot(ep2$sdev, type="b",xlab="Principal component",ylab="Variance",main="Echo 2")

plot(ep1$x[,1],type="l",xlab="Time",ylab="Rotated Data",main="Echo 1")
lines(ep1$x[,2],col="red")
lines(ep1$x[,3],col="blue")
lines(ep1$x[,4],col="green")

plot(ep2$x[,1],type="l",xlab="Time",ylab="Rotated Data",main="Echo 2")
lines(ep2$x[,2],col="red")
lines(ep2$x[,3],col="blue")
lines(ep2$x[,4],col="green")

```

# Principal component analysis, strain rate, actual data

```{r Principal component analysis, strain rate}

sr1 <- echo1$strain
sr2 <- echo2$strain

ep1 <- prcomp(echo1$strain.rate, scale=TRUE) # 70% by pc1
ep2 <- prcomp(echo2$strain.rate, scale=TRUE) # 70% by pc1

ep1 <- prcomp(echo1$strain.rate, scale=FALSE) # 70% by pc1
ep2 <- prcomp(echo2$strain.rate, scale=FALSE) # 70% by pc1

#summary(ep1)
#summary(ep2)

#plot(ep1$rotation[,1], ep1$rotation[,2])
#plot(ep2$rotation[,1], ep2$rotation[,2])

plot(ep1$sdev, type="b",xlab="Principal component",ylab="Variance",main="Echo 1")
plot(ep2$sdev, type="b",xlab="Principal component",ylab="Variance",main="Echo 2")

plot(ep1$x[,1],type="l",xlab="Time",ylab="Rotated Data",main="Echo 1")
lines(ep1$x[,2],col="red")
lines(ep1$x[,3],col="blue")
lines(ep1$x[,4],col="green")

plot(ep2$x[,1],type="l",xlab="Time",ylab="Rotated Data",main="Echo 2")
lines(ep2$x[,2],col="red")
lines(ep2$x[,3],col="blue")
lines(ep2$x[,4],col="green")

```

```{r, Computing principal components with randomly generated data, echo=FALSE,eval=FALSE}

m1 <- (t(apply(samp1, c(1,2), mean)))
plot(1:nrow(m1), m1[,1], type="l",ylim=c(min(m1),max(m1)),xlab="Time Index")
for(i in 1:ncol(m1)){
  lines(1:nrow(m1), m1[,i], col=i)
  }

pr1 <- princomp(m1)
plot(pr1$scores[,1],type="l")
lines(pr1$scores[,2],type="l",col="red")
lines(pr1$scores[,3],type="l",col="blue")
lines(pr1$scores[,4],type="l",col="green")

m2 <- (t(apply(samp2, c(1,2), mean)))
plot(1:nrow(m2), m2[,1], type="l",ylim=c(min(m2),max(m2)))
for(i in 1:ncol(m2)){
  lines(1:nrow(m2), m2[,i], col=i)
  }

pr2 <- princomp(m2)
plot(pr2$scores[,1],type="l")
lines(pr2$scores[,2],type="l",col="red")
lines(pr2$scores[,3],type="l",col="blue")
lines(pr2$scores[,4],type="l",col="green")


```
```{r Calculation of generalized variances, echo=FALSE, eval=FALSE}

m1 <- (t(apply(samp1, c(2,3), var)))

plot(1:nrow(m1), m1[,1], type="l",ylim=c(min(m1),max(m1)))
for(i in 1:ncol(m1)){
  lines(1:nrow(m1), m1[,i], col=i)
  }

m2 <- (t(apply(samp2, c(2,3), var)))

plot(1:nrow(m2), m2[,1], type="l",ylim=c(min(m2),max(m2)))
for(i in 1:ncol(m2)){
  lines(1:nrow(m2), m2[,i], col=i)
  }

trace_store1 <- numeric()
trace_store2 <- numeric()
for(i in 1:nSamples){
  trace_store1[i] <- tr(cov(t(samp1[,,i])))
  trace_store2[i] <- tr(cov(t(samp2[,,i])))
}

det_store1 <- numeric()
det_store2 <- numeric()
for(i in 1:nSamples){
  det_store1[i] <- det(cov(t(samp1[,,i])) ) 
  det_store2[i] <- det(cov(t(samp2[,,i])) )
}

```
```{r canonical polyadic decomposition, eval=FALSE, echo=FALSE}

cp1.1 <- cp(tnsr1, num_components=1)
cp1.2 <- cp(tnsr1, num_components=2)
cp1.3 <- cp(tnsr1, num_components=3)
cp1.10 <- cp(tnsr1, num_components=10)
cp1.100 <- cp(tnsr1, num_components=100)

cp2.1 <- cp(tnsr2, num_components=1)
cp2.2 <- cp(tnsr2, num_components=2)
cp2.3 <- cp(tnsr2, num_components=3)
cp2.10 <- cp(tnsr2, num_components=10)
cp2.100 <- cp(tnsr2, num_components=100)

#cp2 <- cp(samp1)

```
```{r Higher order singular value decomposition, eval=FALSE, echo=FALSE}

hosvd1 <- hosvd(tnsr1)
hosvd2 <- hosvd(tnsr2)

str(tnsr1)
str(tnsr2)

```

## Compute total variances

```{r Compute total variances}

# Total variances, actual data
tr(cov(st1))
tr(cov(st2))

# Total variances, simulated data
totvar1 <- numeric()
totvar2 <- numeric()

for(i in 1:nSamples){
  totvar1[i] <- tr(cov(t(samp1[,,i])))
  totvar2[i] <- tr(cov(t(samp2[,,i])))
  }

boxplot(totvar1, totvar2,names=c("Sample  ","Sample 2"),ylab="Total Variance")

```

# RMSD and RMSF

## Root mean square deviation function

```{r Root mean square deviation}

rmsd <- function(positions){
    # Input must be data frame with rows=times 
    # and columns=probe values (strains, strain rates, etc.)
  
    outputdf <- numeric(length=dim(positions)[1])
    
    for(i in 1:nrow(positions)){
      outputdf[i] <- sqrt(sum( (positions[i,] - positions[1,])^2 ) / length(positions[i,]))
    }
    return(rmsd=outputdf)
}

rmsd1 <- rmsd(echo1$strain)
rmsd2 <- rmsd(echo2$strain)

plot(rmsd1,type="l",xlab="Time",ylab="RMS Deviation", main="Echo 1 RMSD")
plot(rmsd2,type="l",xlab="Time",ylab="RMS Deviation", main="Echo 2 RMSD")

```

## Root mean square fluctuation function

```{r Root mean square fluctuation}

# RMSF is taken as average per particle, not per time (which is RMSD)
rmsf <- function(positions){
    # Input must be data frame with rows=times 
    # and columns=probe values (strains, strain rates, etc.)
  
    outputdf <- numeric(length=dim(positions)[2])
    
    for(i in 1:ncol(positions)){
      outputdf[i] <- sqrt(sum( (positions[,i] - positions[,1])^2 ) / length(positions[,i]))
    }
    return(rmsf=outputdf)
}

rmsf1 <- rmsf(echo1$strain)
rmsf2 <- rmsf(echo2$strain)

plot(rmsf1,type="l",xlab="Position Index",ylab="RMS Fluctuation",main="Echo 1 RMSF")
plot(rmsf2,type="l",xlab="Position Index",ylab="RMS Fluctuation",main="Echo 2 RMSF")

```

## Generation of random RMSD sequences

Model the sample RMSDs as ARIMA processes, and simulate from those ARIMA models.

## Function to generate random RMSDs

```{r Generate random RMSDs}
library(forecast)

m2 <- auto.arima(rmsd2)

simulate_RMSDs <- function(RMSDvec, n=100){

  m1 <- auto.arima(RMSDvec)
  nSimRMSD <- n; RMSDmat <- matrix(data=NA, nrow=length(RMSDvec), ncol=nSimRMSD)

  for(i in 1:nSimRMSD){
    RMSDmat[,i] <-  simulate.Arima(m1) }
  
  return(RMSDmat)
  }

```

## Plot RMSD simulationsa

This is an example of the randomly generated sequences.

```{r Random RMSD plot}
randRMSDs1 <- simulate_RMSDs(rmsd1,n=10)
randRMSDs2 <- simulate_RMSDs(rmsd2, n=10)

plot(randRMSDs1[,1],type="l",ylim=c(min(randRMSDs1), max(randRMSDs1) ),xlab="Time Index",ylab="RMSD",main="Randomly Generated RMSDs")

for(i in 1:dim(randRMSDs1)[2]){ lines(randRMSDs1[,i],type="l", col=i) }
```

## Simulated RMSD trajectory statistics

```{r Simulated RMSD trajectory statistics}

# Total variance of RMSDs
tr(cov(randRMSDs1))
tr(cov(randRMSDs2))

# Generalized Variance of RMSDs
det(cov(randRMSDs1))
det(cov(randRMSDs2))

eigs1 <- eigen(cov(randRMSDs1))
eigs2 <- eigen(cov(randRMSDs2))

plot(eigs1$values, type="b",ylab="Eigenvalue",xlab="Eigenvalue Index",main="Echo 1 Simulations")
plot(eigs2$values, type="b",ylab="Eigenvalue",xlab="Eigenvalue Index",main="Echo 2 Simulations")

plot(eigs1$vectors, xlab="PC1", ylab="PC2",main="Eigenvectors of RMSD Trajectories, Sim. 1")
plot(eigs2$vectors, xlab="PC1", ylab="PC2",main="Eigenvectors of RMSD Trajectories, Sim. 2")


cor(eigs1$vectors)
```


# Pilot Data

## Function to read in pilot data (40 total)

```{r Read in four sets of pilot data}
obs1_4ch_path <- "/Users/kwj/Dropbox/LV_Strain/data/pilot_data/obs1/4ch"
obs2_4ch_path <- "/Users/kwj/Dropbox/LV_Strain/data/pilot_data/obs2/4ch"

obs1_sax_path <- "/Users/kwj/Dropbox/LV_Strain/data/pilot_data/obs1/Sax"
obs2_sax_path <- "/Users/kwj/Dropbox/LV_Strain/data/pilot_data/obs2/Sax"

read_pilot_data <- function(path_to_data){
  filenames <- dir(path_to_data)
  nFiles <- length(filenames)
  results <- vector("list")
  
  for(file in 1:length(filenames)){
    results[[ eval(filenames[file]) ]] <- read.echo(paste0(path_to_data, "/", filenames[[file]]))
  }

  return(results)
}

# Return four lists, each containing sublists
obs1_4ch <- read_pilot_data(obs1_4ch_path)
obs2_4ch <- read_pilot_data(obs2_4ch_path)
obs1_sax <- read_pilot_data(obs1_sax_path)
obs2_sax <- read_pilot_data(obs2_sax_path)

```

## Calculate RMSDs of pilot data (4 chamber)

```{r}


extract_rmsd <- function(biglist, sub_index=3){
  # Sub index:
  # 1 = frame.time # 2 = time.progression
  # 3 = strain # 4 = strain.rate # 5 = radial.strain
  # 6 = radial.strain.rate # 7 = volume # 8 = seg.vol
  
  filenames <- names(biglist)
  nEcho <- length(biglist)
  rmsd_storage <- vector("list")
  
  for(i in 1:nEcho){
  rmsd_results <- rmsd( biglist[[i]][[sub_index]] ) 
  rmsd_storage[[filenames[i]]] <- rmsd_results
  }
  
  return(rmsd_storage)
}
  
obs1_4ch_rmsd <- extract_rmsd(obs1_4ch)
obs2_4ch_rmsd <- extract_rmsd(obs2_4ch)
obs1_sax_rmsd <- extract_rmsd(obs1_sax)
obs2_sax_rmsd <- extract_rmsd(obs2_sax)

```

